---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "README-"
)
```

# valaddin

[![Travis-CI Build Status](https://travis-ci.org/egnha/valaddin.svg?branch=master)](https://travis-ci.org/egnha/valaddin)
[![codecov](https://codecov.io/gh/egnha/valaddin/branch/master/graph/badge.svg)](https://codecov.io/gh/egnha/valaddin)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/valaddin)](https://cran.r-project.org/package=valaddin)
 [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

Dealing with invalid function inputs is a chronic pain for R users, given R's 
weakly typed nature. _valaddin_ provides pain relief—a lightweight R package 
that enables you to transform an existing function into a function with input 
validation checks, _in situ_, in a manner suitable for both programmatic use and
interactive sessions.

## Installation

Install from [CRAN](https://cran.r-project.org/package=valaddin)

```{r, eval = FALSE}
install.packages("valaddin")
```

or get the development version from GitHub using the
[devtools](https://github.com/hadley/devtools) package

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("egnha/valaddin", build_vignettes = TRUE)
```

## Why use valaddin

### Fail fast—save time, spare confusion

You can be more confident your function works correctly, when you know its 
arguments are well-behaved. But when they aren't, its better to stop immediately
and bring them into line, than to let them pass and wreak havoc, exposing 
yourself to breakages or, worse, silently incorrect results. Validating the 
inputs of your functions is good [defensive 
programming](http://adv-r.had.co.nz/Exceptions-Debugging.html#defensive-programming)
practice.

Suppose you have a function `secant()`

```{r}
secant <- function(f, x, dx) (f(x + dx) - f(x)) / dx
```

and you want to ensure that the user (or some code) supplies numerical inputs 
for `x` and `dx`. Typically, you'd rewrite `secant()` so that it stops if this 
condition is violated:

```{r, error = TRUE, purl = FALSE}
secant_numeric <- function(f, x, dx) {
  stopifnot(is.numeric(x), is.numeric(dx))
  secant(f, x, dx)
}

secant_numeric(log, 1, .1)

secant_numeric(log, "1", ".1")
```

### The standard approach in R is problematic

While this works, it's not ideal, even in this simple situation, because

* it's inconvenient for interactive use at the console: you have to declare a 
new function, and give it a new name (or copy-paste the original function body)

* it doesn't catch all errors, only the first that occurs among the checks

* you're back to square one, if you later realize you need additional checks, or
want to skip them altogether.

### valaddin rectifies these shortcomings

valaddin provides a function `firmly()` that takes care of input validation by
_transforming_ the existing function, instead of forcing you to write a new one.
It also helps you by reporting _every_ failing check.

```{r, error = TRUE, purl = FALSE}
library(valaddin)

# Check that `x` and `dx` are numeric
secant <- firmly(secant, list(~x, ~dx) ~ is.numeric)

secant(log, 1, .1)

secant(log, "1", ".1")
```

To add additional checks, just apply the same procedure again:

```{r, error = TRUE, purl = FALSE}
secant <- firmly(secant, list(~x, ~dx) ~ {length(.) == 1L})

secant(log, "1", c(.1, .01))
```

Or, alternatively, all in one go:

```{r, error = TRUE, purl = FALSE}
secant <- loosely(secant)  # Retrieves the original function
secant <- firmly(secant, list(~x, ~dx) ~ {is.numeric(.) && length(.) == 1L})

secant(log, 1, .1)

secant(log, "1", c(.1, .01))
```

### Check anything using a simple, consistent syntax

`firmly()` uses a simple formula syntax to specify arbitrary checks—not just 
type checks. Every check is a formula of the form `<where to check> ~ <what to 
check>`. The "what" part on the right is a _function_ that does a check, while 
the (form of the) "where" part on the left indicates where to apply the check—at
which _arguments_ or _expressions_ thereof.

valaddin provides a number of conveniences to make checks for `firmly()` 
informative and easy to specify.

#### Use custom error messages

Use a custom error message to clarify the _purpose_ of a check:

```{r, error = TRUE, purl = FALSE}
bc <- function(x, y) c(x, y, 1 - x - y)

# Check that `y` is positive
bc_uhp <- firmly(bc, list("(x, y) not in upper half-plane" ~ y) ~ {. > 0})

bc_uhp(.5, .2)

bc_uhp(.5, -.2)
```

#### Easily apply a check to all arguments

Leave the left-hand side of a check formula blank to apply it to all arguments:

```{r, error = TRUE, purl = FALSE}
bc_num <- firmly(bc, ~is.numeric)

bc_num(.5, ".2")

bc_num(".5", ".2")
```

Or fill in a custom error message:

```{r, error = TRUE, purl = FALSE}
bc_num <- firmly(bc, "Not numeric" ~ is.numeric)

bc_num(.5, ".2")
```

#### Check conditions with multi-argument dependencies

Use the `isTRUE()` predicate to implement checks depending on multiple arguments
or, equivalently, the check maker `vld_true()`:

```{r, error = TRUE, purl = FALSE}
in_triangle <- function(x, y) {x >= 0 && y >= 0 && 1 - x - y >= 0}
outside <- "(x, y) not in triangle"

bc_tri <- firmly(bc, list(outside ~ in_triangle(x, y)) ~ isTRUE)

# Or more concisely:
bc_tri <- firmly(bc, vld_true(outside ~ in_triangle(x, y)))

# Or more concisely still, by relying on an auto-generated error message:
# bc_tri <- firmly(bc, vld_true(~in_triangle(x, y)))

bc_tri(.5, .2)

bc_tri(.5, .6)
```

Alternatively, use the `lift()` function from the 
[purrr](https://github.com/hadley/purrr) package:

```{r, message = FALSE}
library(purrr)

bc_tri <- firmly(bc, list(outside ~ list(x, y)) ~  purrr::lift(in_triangle))
```

#### Layer checks using the magrittr pipe `%>%`

Activate checks in stages using the
[magrittr](https://github.com/tidyverse/magrittr) pipe `%>%`:

```{r, error = TRUE, message = FALSE, purl = FALSE}
library(magrittr)

bc <- {
  function(x, y) c(x, y, 1 - x - y)
} %>%
  firmly("Not numeric" ~ is.numeric, "Not scalar" ~ {length(.) == 1L}) %>%
  firmly(vld_true(outside ~ in_triangle(x, y)))
                   
bc(.5, .2)

bc(.5, c(.2, .1))

bc(".5", 1)
```

### Learn more

See the package documentation `?firmly`, `help(p = valaddin)` for detailed 
information about `firmly()` and its companion functions, and the vignette 
`vignette("valaddin")` for an overview of use cases.

## Related packages

* [assertive](https://bitbucket.org/richierocks/assertive), 
[assertthat](https://github.com/hadley/assertthat), and 
[checkmate](https://github.com/mllg/checkmate) provide handy collections of 
predicate functions that you can use in conjunction with `firmly()`.

* [argufy](https://github.com/gaborcsardi/argufy) takes a different approach to 
input validation, using [roxygen](https://github.com/klutometis/roxygen) 
comments to specify checks.

* [ensurer](https://github.com/smbache/ensurer) and 
[assertr](https://github.com/ropensci/assertr) provide a means of validating 
function values. Additionally, ensurer provides an experimental replacement for 
`function()` that builds functions with type-validated arguments.

* [typeCheck](https://github.com/jimhester/typeCheck), together with [Types for 
R](https://github.com/jimhester/types), enables the creation of functions with 
type-validated arguments by means of special type annotations. This approach is 
orthogonal to that of valaddin: whereas valaddin specifies input checks as 
_predicate functions with scope_, typeCheck specifies input checks as _arguments
with type_.

## License

MIT Copyright © 2017 [Eugene Ha](https://github.com/egnha)
